A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-09-08 docs/discussion_ML_v1.02.tex
Started writing discussion of possible minor merger.

All of this leaves the central dip unexplained: It cannot be due to tangential velocity anisotropy in the center alone (see Section \ref{sec:results_JAM_SB}, Figure \ref{fig:JAM_modelA2}). It also cannot be explained by a strong contribution of dark matter inside the bulge together with a moderate tagential velocity anisotropy, because the corresponding dark matter halos would still be too massive to fit the data in the outer regions (see Section \ref{sec:results_JAM_NFW}, Figure \ref{fig:modelB4_vrms}]).

@@

The above discussion motivates the following speculation: In the absence of a strong dark matter contribution in the center, the overall stellar-mass-to-light ratio within the Einstein radius indicates a bottom-heavy IMF close to the Salpeter IMF, consistent with estimates from the velocity dispersion. At the same time the central dip can then be only explained, if there was an increase in stellar mass-to-light ratio with radius \textit{within} the bulge. \emph{If} there was a central stellar population with an IMF close to the \citet{Chabrier2003} IMF, surrounded by a more bottom-heavy population, the central stellar kinematics would be well explained and be fully consistent with the lensing results. (According to Figure \ref{fig:lenscompareboth} the lensing result might not predict a mass to light ratio gradient inside $R_\text{ein}$, but then again the mass slope $alpha=1$ was only weakly constrained.) It is not expected that the bluish disk of J1331 has a $\Upsilon_\text{I,*}$ as high as the reddish bulge. The kinematics at larger radii would then also be much better recovered by lower mass dark matter halo as found by \citet{SWELLSV}. \\

Standard JAM modelling approaches seem not to work for J1331. A JAM model for J1331 would need to allow for stellar mass-to-light ratio gradients within the galaxy, velocity anisotropy and a dark matter halo. Because of degeneracies between stellar and dark mass, and matter distribution and anisotropy profile, such a dynamical model would not lead to very tight constraints on the model parameters. Further priors are needed. Detailed stellar population analyses of the spectra taken along J1331's major axis could possibly support or contradict the suspicion of the existence of stellar mass-to-light ratio gradients in J1331.

>>>>

All of this leaves the central dip unexplained: It cannot be due to tangential velocity anisotropy in the center alone (see Section \ref{sec:results_JAM_SB}, Figure \ref{fig:JAM_modelA2}). It also cannot be explained by a strong contribution of dark matter inside the bulge together with a moderate tagential velocity anisotropy, because the corresponding dark matter halos would still be too massive to fit the data in the outer regions (see Section \ref{sec:results_JAM_NFW}, Figure \ref{fig:modelB4_vrms}]).\\

@@

Another feature in the stellar kinematics, that none of our JAM models was able to reproduce in the slightest and which we therefore excluded in the modelling, was the dip in $v_\text{rms}$ around $\sim 6$ arcsec, which occurs around the transition from bulge to disk (see Figure [TO DO: 1b]).\\

The above discussion motivates the following speculation: In the absence of a strong dark matter contribution in the center, the overall stellar-mass-to-light ratio within the Einstein radius indicates a bottom-heavy IMF close to the Salpeter IMF, consistent with estimates from the velocity dispersion. At the same time the central dip can then be only explained, if there was an increase in stellar mass-to-light ratio with radius \textit{within} the bulge. \emph{If} there was a central stellar population with an IMF close to the \citet{Chabrier2003} IMF, surrounded by a more bottom-heavy population, the central stellar kinematics would be well explained and be fully consistent with the lensing results. (According to Figure \ref{fig:lenscompareboth} the lensing result might not predict a mass to light ratio gradient inside $R_\text{ein}$, but then again the mass slope $alpha=1$ was only weakly constrained.) The disk of J1331 has a lower $\Upsilon_\text{I,*}$ than the bulge (see Table \ref{tab:previousresults} according to \citet{SWELLSI}). Such a drop in $\Upsilon_\text{I,*}$ at the transition region from bulge to disk around $\sim 5$ arcsec could lead to the observed drop in $v_\text{rms}$, while an increasing contribution of a lower mass dark matter halo at larger radii as found by \citet{SWELLSV}, would recover the kinematics in the out regions of J1331's disk.\\

### 2015-09-08 docs/discussion_merger_v1.01.tex
Started writing discussion of possible minor merger.

[TO DO]

>>>>

\paragraph{TO DO: Stuff to mention}
\begin{itemize}
\item J1331 has a counter-rotating stellar component.
\item suggests that there was a process where two components with angular momenta oriented in opposite directions were involved
\item 1. scenario: accretion of gas on retrograde orbits and subsequent star formation
\item --> counter-rotating component is younger than the rest of the galaxy
\item argument against: To form enough stars such taht the net rotation of the large and massive core is retrograde, a very large amount of gas would have had to be accreted - which is not very likeliy [TO DO" REF: KLEINEBERG 2011] 
\item 2. scenarios: Major mergers of similarly massive galaxies
\item argument against: major merger products are often quiescent, elliptical galaxies [TO DO: REF: TOOMRE TOOMRE 1972]
\item 3. sencaro: minor merger
\item --> dissipationless accretion of a satellite on retrograde orbit [TO DO: REF: KORMENDY 1984] with dense core to survive and spiral to core due to tidal friction
\item --> counter-rotating component would be younger, as star formation in low-mass satellite galaxies occurs later [TO DO: REF]
\item if the counter-rotating component is younger needs to be tested by detailed stellar population analysis
\item Minor merger effects in elliptical galaxies have been studied e.g. by [TO DO:REF: BALCELLS AND QUINN 1990]
\item large bulges of sporals or ellipticals are reddest in the center and become bluer further out
\item Satellite nucleus settling in the core of the larger galaxy could caus a reverse color gradient
\item Normally this gradient is only weak, but there are exceptions [TO DO: CAROLLO 1997].
\item A second effect of the minor merger could be a misalignement of the photometric and kinematic major axis
\end{itemize}

\subsection{Future Work}

Standard JAM modelling approaches seem not to work for J1331. A JAM model for J1331 would need to allow for stellar mass-to-light ratio gradients within the galaxy, velocity anisotropy and a dark matter halo. Because of degeneracies between stellar and dark mass, and matter distribution and anisotropy profile, such a dynamical model would not lead to very tight constraints on the model parameters. Further priors are needed. Detailed stellar population analyses of the spectra taken along J1331's major axis could possibly support or contradict the suspicion of the existence of stellar mass-to-light ratio gradients in J1331.

\paragraph{Model Failures}
\begin{itemize}
\item Assumption of axisymmetry is not true anymore
\item [TO DO: more]
\end{itemize}

### 2015-09-08 docs/model_JAM_v1.01.tex
Started writing discussion of possible minor merger.

\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also \citet{1994A&A...285..723E}). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Eq. (\ref{eq:deprojMGE}). The mass density $\rho(R,z)$ consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. (To mimick gradients of mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$.) To add a Navarro-Frenck-White (NFW) \citep{Navarro+1995c,NFW96} dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: Maybe remove detailed description of the mass MGE here, if it is described in detail somewhere else.] The gravitational potential is generated from the mass MGE by integrating the Poisson equation \citep{1994A&A...285..723E}. Equations \ref{eq:Jeans1} and \ref{eq:Jeans2} together with \ref{eq:bz} provide the velocity dispersion tensor $\langle v_i v_j \rangle = \langle v_i^2 \rangle$ (with $i,j \in \{ R,\phi, z\}$). It is then rotated by the inclination angle $i$ to the coordinate system of the observer ($(x',y')$ being the plane of the sky and $z'$ the line-of-sight, where $x'$ is alinged with the galaxy's major axis). Taking a light-weighted projection along the line-of-sight gives a model prediction for the line-of-sight velocity second moment, which is comparable to actual spectroscopic measurements of the second velocity moment. Details of the derivation using the MGE formalism are given in \citet{Cap08} and repeated in \citet{GlennEC}, we therefore just quote the result by \citet{Cap08} (their Equation 28) for the line-of-sight second velocity moment prediction from the Jeans equations,

@@

&&\left(I \langle v_\text{los}^2\rangle\right)(x',y')\\

@@

&& \times \frac{\sigma_k^2 \ q^{2}_k \left( \cos^2 i + \frac{\sin^2 i}{1-\beta_{z,k}}\right) + \mathscr{D} \  {x'}^{2} \sin^2 i}{(1-\mathscr{C}u^2) \sqrt{(\mathscr{A} + \mathscr{B} \cos^2 i) \left[1-(1-q^{2}_j) u^2 \right]}}\label{eq:explicitLOSvelCap}\\
&& \times \exp\left\{- \mathscr{A} \left[{x'}^{2} + \frac{(\mathscr{A}+\mathscr{B}) \ {y'}^{2}}{\mathscr{A}+\mathscr{B}\cos^2 i}\right] \right\} \diff u, \\

@@

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ [TO DO: consistent, los or rot] and velocity dispersions $\sigma$ as described in \S\ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $v_\text{los}$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Eq. [TO DO] predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified to model the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1$ arcsec \citep{SWELLSV} and a certain given extent in along the major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x,y) = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x \leq + \frac{L_x}{2} \text{ and } - \frac{L_y}{2} \leq y \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise} \end{array} \right. .
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian 
\begin{equation*}
\text{PSF}(x,y)=\mathscr{N}(0,FWHM/2\sqrt{2\ln2})
\end{equation*}
with FWHM=1.1 arcsec  \citep{SWELLSV}. The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x,y) &=& (\text{PSF} \ast \text{AP})(x,y) \\
&=& \frac{1}{4} \prod_{u \in \{x,y\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x,y)$ [TO DO: primed x and y or not????]
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with with the convolution kernel, the JAM code by \citet{Cap08} [TO DO: reference code] performs the convolution numerically. We set $L_x = 0.21$ arcsec as the width of the model pixel, and get a prediction for the actual measurements in bins of width 0.63, 1.26 and 1.89 arcsec \citep{SWELLSV} as light-weighted mean from each 3, 6 and 9 model pixels.

\paragraph{Rotation curve.} The intrinsic rotation curve is the first velocity moment $\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2}$. The observed rotation curve is the projection of the light-weighted contributions to $\langle v_\phi\rangle$ along the line-of-sight(\cite{Cap08}),
\begin{equation*}
I \langle v_\text{los}\rangle = \int_{-\infty}^{+\infty} \nu \langle v_\phi\rangle \cos \phi \sin i \diff z'.
\end{equation*}
The first velocity moments cannot be uniquely determined from the Jeans equations, which give only a prediction for the second velocity moments. Further assumptions are needed to separate the second velocity moments into ordered and random motion. \citet{Cap08} assumes that in a steady state there is no streaming velocity in $R$ direction, i.e. $\langle v_R \rangle = 0$ and therefore $\sigma_R^2 = \langle v_R^2 \rangle$. Then \citet{Cap08} relates the dispersions in $R$ and $\phi$ direction such that
$$\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2} \equiv \kappa \sqrt{\langle v_\phi^2 \rangle - \langle v_R^2 \rangle},$$
and the $\kappa$ parameter quantifies the rotation: $\kappa = 0$ means no rotation at all and $|\kappa| = 1$ describes a velocity dispersion ellipsoid that is a circle in the $R$-$\phi$ plane \citep{Cap08}. The sign of $\kappa$ determines the rotation direction. We can assign a constant $\kappa_k$ to every Gaussian in the MGE formalism and
\begin{equation*}
\nu \langle v_\phi\rangle = \left[\nu \sum_{k} \kappa_k^2 \left( [\nu\langle v_\phi^2 \rangle]_k - [\nu\langle v_R^2 \rangle]_k\right) \right]^{1/2}
\end{equation*} 
is then the light-weighted circular velocity curve, given the second velocity moments found from the Jeans equations.
\\To model the counter-rotating core of J1331 with one free parameter for, we employ the condition that the overall $\kappa(R)$ profile should smoothly and relatively steeply transition from $\kappa(R) = -\kappa' < 0$ at small $R$ [TO DO: Check, ich glaube das hier ist das intrinsische R in Zylinder-Coordinaten] trough $\kappa(R_0) \equiv 0$ and increase to $\kappa(R) = \kappa' > 0$ at large $R$. Our imposed profile is
\begin{equation}
\kappa(R) = \kappa' \frac{R^2 - R_0^2}{R^2 + R_0^2}. \label{eq:kappa_profile}
\end{equation}
We find $\kappa'$ by matching the model $\langle v_\text{los} \rangle$ with the symmetrized $v_\text{rot}$ data, where for a given $\kappa'$ the $\kappa_k$ are found from fitting the MGE generated profile $\kappa(R) = \sum_k \kappa_k \nu_k(r)/\sum_k \nu_k(r)$ to Eq. (\ref{eq:kappa_profile}). The observed zero point is at $R'_0\approx 2$ arcsec. In the deprojected galactic plane the radius of zero rotation would be at a $R_0 \gtrsim 2$ arcsec, and we choose it to be at 2.2 arcsec.

>>>>

\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also \citet{1994A&A...285..723E}). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Equation \ref{eq:deprojMGE}. The mass density $\rho(R,z)$ consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. (To mimick gradients of mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$.) To add a Navarro-Frenck-White (NFW) \citep{Navarro+1995c,NFW96} dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: Maybe remove detailed description of the mass MGE here, if it is described in detail somewhere else.] The gravitational potential is generated from the mass MGE by integrating the Poisson equation \citep{1994A&A...285..723E}. Equations \ref{eq:Jeans1} and \ref{eq:Jeans2} together with \ref{eq:bz} provide the velocity dispersion tensor $\langle v_i v_j \rangle = \langle v_i^2 \rangle$ (with $i,j \in \{ R,\phi, z\}$). It is then rotated by the inclination angle $i$ to the coordinate system of the observer ($(x',y')$ being the plane of the sky and $z'$ the line-of-sight, where $x'$ is alinged with the galaxy's major axis). Taking a light-weighted projection along the line-of-sight gives a model prediction for the line-of-sight velocity second moment, which is comparable to actual spectroscopic measurements of the second velocity moment. Details of the derivation using the MGE formalism are given in \citet{Cap08} and repeated in \citet{GlennEC}, we therefore just quote the result by \citet{Cap08} (their Equation 28) for the line-of-sight second velocity moment prediction from the Jeans equations,

@@

&&\left(I \langle v_\text{los}^2\rangle\right)(x',y')\nonumber\\

@@

&& \times \frac{\sigma_k^2 \ q^{2}_k \left( \cos^2 i + \frac{\sin^2 i}{1-\beta_{z,k}}\right) + \mathscr{D} \  {x'}^{2} \sin^2 i}{(1-\mathscr{C}u^2) \sqrt{(\mathscr{A} + \mathscr{B} \cos^2 i) \left[1-(1-q^{2}_j) u^2 \right]}}\nonumber\\
&& \times \exp\left\{- \mathscr{A} \left[{x'}^{2} + \frac{(\mathscr{A}+\mathscr{B}) \ {y'}^{2}}{\mathscr{A}+\mathscr{B}\cos^2 i}\right] \right\} \diff u, \label{eq:explicitLOSvelCap}

@@

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ [TO DO: consistent, los or rot] and velocity dispersions $\sigma$ as described in Section \ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $\langle v_\text{los}^2 \rangle$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Equation \ref{eq:explicitLOSvelCap} predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified according to the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1$ arcsec \citep{SWELLSV} and a certain given extent in along the galaxy's major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x,y) = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x \leq + \frac{L_x}{2}\\
& \text{and } - \frac{L_y}{2} \leq y \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise.} \end{array} \right.
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian with Full Width Half Maximum (FWHM) of1.1 arcsec \citep{SWELLSV}
\begin{equation*}
\text{PSF}(x,y)=\mathscr{N}(0,\text{FWHM}/2\sqrt{2\ln2}).
\end{equation*}
The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x,y) &=& (\text{PSF} \ast \text{AP})(x,y) \\
&=& \frac{1}{4} \prod_{u \in \{x,y\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x,y)$ [TO DO: primed x and y or not????]
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with the convolution kernel, the JAM code by \citet{Cap08} [TO DO: reference code] performs the convolution numerically. We set $L_x = 0.21$ arcsec as the width of the model pixel, and get a prediction for the actual measurements in bins of width 0.63, 1.26 and 1.89 arcsec \citep{SWELLSV} as light-weighted mean from each 3, 6 and 9 model pixels.

\paragraph{Rotation curve.} The intrinsic rotation curve is the first velocity moment $\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2}$. The observed rotation curve is the projection of the light-weighted contributions to $\langle v_\phi\rangle$ along the line-of-sight(\cite{Cap08}),
\begin{equation*}
I \langle v_\text{los}\rangle = \int_{-\infty}^{+\infty} \nu \langle v_\phi\rangle \cos \phi \sin i \diff z'.
\end{equation*}
The first velocity moments cannot be uniquely determined from the Jeans equations, which give only a prediction for the second velocity moments. Further assumptions are needed to separate the second velocity moments into ordered and random motion. \citet{Cap08} assumes that in a steady state there is no streaming velocity in $R$ direction, i.e. $\langle v_R \rangle = 0$ and therefore $\sigma_R^2 = \langle v_R^2 \rangle$. Then \citet{Cap08} relates the dispersions in $R$ and $\phi$ direction such that
$$\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2} \equiv \kappa \sqrt{\langle v_\phi^2 \rangle - \langle v_R^2 \rangle},$$
and the $\kappa$ parameter quantifies the rotation: $\kappa = 0$ means no rotation at all and $|\kappa| = 1$ describes a velocity dispersion ellipsoid that is a circle in the $R$-$\phi$ plane \citep{Cap08}. The sign of $\kappa$ determines the rotation direction. We can assign a constant $\kappa_k$ to every Gaussian in the MGE formalism and
\begin{equation*}
\nu \langle v_\phi\rangle = \left[\nu \sum_{k} \kappa_k^2 \left( [\nu\langle v_\phi^2 \rangle]_k - [\nu\langle v_R^2 \rangle]_k\right) \right]^{1/2}
\end{equation*} 
is then the light-weighted circular velocity curve, given the second velocity moments found from the Jeans equations.
\\To model the counter-rotating core of J1331 with one free parameter for, we employ the condition that the overall $\kappa(R)$ profile should smoothly and relatively steeply transition from $\kappa(R) = -\kappa' < 0$ at small $R$ [TO DO: Check, ich glaube das hier ist das intrinsische R in Zylinder-Coordinaten] trough $\kappa(R_0) \equiv 0$ and increase to $\kappa(R) = \kappa' > 0$ at large $R$. Our imposed profile is
\begin{equation}
\kappa(R) = \kappa' \frac{R^2 - R_0^2}{R^2 + R_0^2}. \label{eq:kappa_profile}
\end{equation}
We find $\kappa'$ by matching the model $\langle v_\text{los} \rangle$ with the symmetrized $v_\text{rot}$ data, where for a given $\kappa'$ the $\kappa_k$ are found from fitting the MGE generated profile $\kappa(R) = \sum_k \kappa_k \nu_k(r)/\sum_k \nu_k(r)$ to Equation \ref{eq:kappa_profile}. The observed zero point is at $R'_0\approx 2$ arcsec. In the deprojected galactic plane the radius of zero rotation would be at a $R_0 \gtrsim 2$ arcsec, and we choose it to be at 2.2 arcsec.
